---
title: "Mixed Modeling for Repeated Measures and Hierarchical Data"
author: "Jocelyn Stalker"
date: "9/23/2021"
output: html_document
---
<body style="background-color:darkolivegreen;">

```{css, include=FALSE}
.Autumn {
  background-color: firebrick3;
}
```
<style>
div.cranberries pre { background-color:#D64D32; }
div.cranberries pre.r { background-color:#EE9B6B; }
</style>

<div class = "cranberries">
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(janitor)
library(tidyverse)
library(dplyr)
library(stringr)
library(lme4)
library(lmerTest)
library(performance)
library(emmeans)
library(RColorBrewer)
library(colorRamps)
```

# Hierarchical Data

```{r import and prepare data, class.source="Autumn"}
hurr.herp <- read.csv("./Data/hurricane_herp_data.csv", header=TRUE, na.strings= "") %>% clean_names() %>% filter(!str_detect(sitio_estadios, "NA")) %>% filter(!str_detect(grupo, "TORTUGA"))
hurr.herp$grupo <- as.factor(hurr.herp$grupo)
levels(hurr.herp$grupo) <- c("amphibian","lizard","lizard","snake","turtle")
hurr.herp$sitio_estadios <- as.factor(hurr.herp$sitio_estadios)
levels(hurr.herp$sitio_estadios) <- c("A","B","C","D","E","F","G","H","I","J","K","L","M","N","O")
hurr.herp <- select(hurr.herp, sitio_estadios, pre_post, grupo, especie)
colnames(hurr.herp) <- c("site","pre_post","group","species")
hurr.herp[469, "species"] <- "Phrynohyas venulosa" # Fix a typo in the data
hurr.herp.group <- hurr.herp %>% count(site, pre_post, group)
hurr.herp.group$pre_post = factor(hurr.herp.group$pre_post, levels = c("Pre", "Post")) # Reorder factor to be pre then post
hurr.herp.snake <- hurr.herp %>% count(site, pre_post, group) %>% filter(str_detect(group, "snake"))
hurr.herp.snake$pre_post = factor(hurr.herp.snake$pre_post, levels = c("Pre", "Post"))
hurr.herp.amph <- hurr.herp %>% count(site, pre_post, group) %>% filter(str_detect(group, "amphibian"))
hurr.herp.amph$pre_post = factor(hurr.herp.amph$pre_post, levels = c("Pre", "Post"))
hurr.herp.liz <- hurr.herp %>% count(site, pre_post, group) %>% filter(str_detect(group, "lizard"))
hurr.herp.liz$pre_post = factor(hurr.herp.liz$pre_post, levels = c("Pre", "Post"))
```

```{r geom_jitter plot, fig.height= 6}
getPalette = colorRampPalette(brewer.pal(9, "Oranges"))

ggplot(hurr.herp.group, aes(group, n, colour= site, shape= pre_post)) + geom_jitter(width =0.15, size=5) + xlab("Taxonomic Group") +
  ylab("Number of Individuals") + scale_color_manual(name= "Site", values= getPalette(15)) + scale_shape_discrete(name= "Time Relative\n to Hurricane") +
  ggtitle("Herpetofauna Counts Before and After a Hurricane")
```

```{r plot the data, class.source="Autumn"}
ggplot(hurr.herp.snake, aes(x=pre_post, y=n)) + xlab("Time Relative to Hurricane") + ylab("Number of Individuals") +
  geom_line(aes(group=site), color= "chocolate2") +
  geom_point(size=3, color= "burlywood4") +
  ggtitle("Effect of a Hurricane on Snake Counts")

ggplot(hurr.herp.liz, aes(x=pre_post, y=n)) + xlab("Time Relative to Hurricane") + ylab("Number of Individuals") +
  geom_line(aes(group=site), color= "chocolate2") +
  geom_point(size=3, color= "burlywood4") +
  ggtitle("Effect of a Hurricane on Lizard Counts")

ggplot(hurr.herp.amph, aes(x=pre_post, y=n)) + xlab("Time Relative to Hurricane") + ylab("Number of Individuals") +
  geom_line(aes(group=site), color= "chocolate2") +
  geom_point(size=3, color= "burlywood4") +
  ggtitle("Effect of a Hurricane on Amphibian Counts")
```

```{r fit mixed model for herps}
lmer.herp <- lmer(n ~ group*pre_post + (1|site), data= hurr.herp.group)
```

```{r check machine model for herps, fig.width=9.5, fig.height=9}
performance::check_model(lmer.herp)
```

```{r run ANOVA on model}
anova(lmer.herp)
```

Let's calculate model-adjusted means.

```{r fit emmeans from model}
emmeans(lmer.herp, "pre_post")
```

Pre-hurricane means are lower than post-hurricane means, but this doesn't tell us much. We know from our ANOVA that this is statistically non-significant.

```{r fit emmeans from model - each group*pre_post}
herp.emm <- emmeans(lmer.herp, "pre_post", "group")
herp.emm
```

```{r emmeans as dataframe}
data.emm <- as.data.frame(summary(herp.emm))
data.emm
```

Model-adjusted means are the same as raw means because the experiment is fully balanced, but look, I'm proving I can run the code for them!

```{r plot pre-post emmeans}
data.emm$group <- factor(data.emm$group, levels= c("lizard","amphibian","snake"))
ggplot(data.emm, aes(x=pre_post, y=emmean, group=group, color=group)) + scale_color_manual(name= "Taxonomic Group", values= c("darkgoldenrod1","burlywood4","chocolate2")) +
  geom_line() +
  geom_point(size=4)+
  geom_errorbar(aes(ymin=emmean-SE, ymax=emmean+SE), width=.2)+
  labs(title="Herpetofauna Counts Before and After a Hurricane", x= "Time Relative to Hurricane", y= "Number of Individuals")
```

# Repeated Measures Data

```{r import and prep data}
chimp <- read.csv("./Data/chimp_personality.csv", header=TRUE) %>% clean_names()
chimp$sex <- as.factor(chimp$sex)
```

```{r plot the chimp data}
ggplot(chimp, aes(x=time, y=extraversion, color=sex)) + scale_x_discrete(labels= c("Test 1", "Test 2")) +
  scale_color_manual(values= c("chocolate2","burlywood4"), name= "Sex", labels= c("Male","Female")) + labs(title= "Change in Extraversion of Chimps", x= "Test Number", y= "Extraversion Rating") + geom_line(aes(group=id)) + geom_point(size=3)
```

```{r fit mixed model}
chimp.lmer <- lmer(extraversion ~ sex*time + (1|id), data= chimp)
anova(chimp.lmer)
summary(chimp.lmer)
```

```{r check machine model, fig.width=9.5, fig.height=9}
performance::check_model(chimp.lmer)
```

```{r emmeans}
chimp.emm <- emmeans(chimp.lmer, "sex")
chimp.emm
chimp.emm.time <- emmeans(chimp.lmer, "time", "sex")
chimp.emm.time
data.chimp.emm <- as.data.frame(chimp.emm.time)
```

Interesting... all chimps, regardless of sex, got less extraverted over time. Males were more extraverted than females.

```{r plotting chimp emmeans}
chimp.means <- chimp %>% group_by(time, sex) %>% summarise(mean_extraversion= mean(extraversion), se_rating =sd(extraversion)/sqrt(n()))
chimp.means

ggplot(data.chimp.emm, aes(x= time, y= emmean, color= sex)) + geom_point(size= 4) +
  geom_errorbar(aes(ymin= emmean-SE, ymax= emmean+SE), width=.2) +
  labs(title= "Means and Corrected Means for Extraversion in Chimps", x= "Test Number", y= "Mean Extraversion Rating") +
  scale_color_manual(values= c("chocolate2","burlywood4"), name= "Sex", labels= c("Male","Female")) +
  scale_x_discrete(labels= c("Test 1","Test 2")) +
  geom_point(data= chimp.means, size=2, x= chimp.means$time, y= chimp.means$mean_extraversion, color="darkgoldenrod1")
```

emmeans are the same as actual means.

</div>